A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-16 latex/main.tex
Add recombinant evidence notebook
Closes #898

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but are not incorporated into UShER trees and must therefore
be accounted for separately.
Because deletions are difficult to distinguish 
from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
(11288+9, 21765+6, 21991+3,
where the first and second numbers are
the start position and length, respectively)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hotspot \cite{McCarthy2021})
for example,
the 21765+6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991+3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis.
The first recombinant lineage XA was designated in 2021
based on the analysis of Jackson et al.\cite{Jackson2021},
which established that recombination with onward transmission had occurred [weak FIXME].
Although the process of incorporating recombinant lineages
in the Pango nomenclature system is well defined,
the identification of such events is much more subjective,
requiring the synthesis of multiple lines of evidence and
often the manual inspection of phylogenies and sequence alignments
by volunteers~\cite{Smith2023tracking}.
% Debatable whether this adds or substracts from narrative
As of pango-designation version v1.33.1 (2025-03-31),
a total of 146 distinct Pango X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146

@@

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
that they are more parsimoniously explained as sublineages or sisters of XQ
than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B, Document \ref{sec:dd_xx}).

@@

We validated the recombination events in the sc2ts ARG by running 
the 3SEQ~\cite{Lam2018improved}, CovRecomb~\cite{Li2024CovRecomb} and rebar 
recombination detection methods on the putative recombinant sequences
(STAR Methods) summarized in Table~\ref{tab:recombinant_validation}.
Overall, validation among the QC-passing recombinants is
the highest for 3SEQ at 88\% and
the lowest for CovRecomb at 25\%.
CovRecomb has been previously shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}, 
and is specifically intended to detect interlineage recombinants.
As expected, CovRecomb missed most of the recombinants that
involve closely related parental Pango lineages
(Figure~\ref{fig:recombinant_evidence_cov_recomb}C),
in contrast to RIPPLES
which can detect intralineage recombinants
(see Figure~\ref{fig:recombinant_evidence_ripples_p4} and
Figure~\ref{fig:recombinant_evidence_ripples_p3}).
The 12\% of recombinants
not passing 3SEQ validation are also the most marginally supported
(see next section; Figure~\ref{fig:recombinant_evidence}).
All of the validation metrics also correlate strongly with the 
QC filter, with very few of the QC-failing recombinants
passing external validation (Table~\ref{tab:recombinant_validation}).

@@

\begin{table}
\caption{Validation of sc2ts recombinants by 3SEQ, CovRecomb and rebar over 
all 855 recombinants (total) and XXX passing and XXX failing QC.
TODO FILL OUT DETAILS.
}
\label{tab:recombinant_validation}
\centering
\begin{tabular}{llll}
\toprule
Method & Overall & QC Pass & QC Fail \\
\midrule
3SEQ & xx & 88\% \\
CovRecomb & \\
rebar & \\
\bottomrule
\end{tabular}
\end{table}

>>>>

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but are not incorporated into UShER trees and must therefore
be accounted for separately.
Because deletions are difficult to distinguish 
from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
(11288:9, 21765:6, 21991:3)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hotspot \cite{McCarthy2021})
for example,
the 21765:6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991:3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis.
The first recombinant lineage XA was designated in 2021
based on the analysis of Jackson et al.\cite{Jackson2021},
which established that recombination with onward transmission had occurred.
Although the process of incorporating recombinant lineages
in the Pango nomenclature system is well defined,
the identification of such events is much more subjective,
requiring the synthesis of multiple lines of evidence and
often the manual inspection of phylogenies and sequence alignments
by volunteers~\cite{Smith2023tracking}.
% Debatable whether this adds or substracts from narrative
As of pango-designation version v1.33.1 (2025-03-31),
a total of 146 distinct Pango X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146

@@

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
% NOTE: deleted the "sister" bit because it's confusing when
% reading the document at this level. Keep the detailed hyptothese
% to the deep dives.
that they are more parsimoniously explained as sublineages of XQ
than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B, Document \ref{sec:dd_xx}).

@@

We validated the recombination events in the sc2ts ARG by running 
3SEQ~\cite{Lam2018improved} on the putative recombinant 
sequence trios (STAR Methods), 
and 383 of the 855 (45\%) were classified as recombinant.
Validation status correlates strongly with our QC filter, with 
88\% of the 354 QC-passing recombinants 
and 14\% of the QC-failing recombinants
being classified as recombinant.
The 12\% of QC-passing recombinants
not validated by 3SEQ are also among the most marginally supported
(see next section; Figure~\ref{fig:recombinant_evidence}).
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar recombination detection methods on the putative recombinant 
sequences, summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate at 25\% 
among the QC-passing recombinants. This is expected, as 
CovRecomb and is specifically intended to detect interlineage recombinants
and been previously shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.

@@

\begin{table}
\caption{Validation of sc2ts recombinants by 3SEQ, CovRecomb and rebar over 
all 855 recombinants and the 354 passing and 501 failing QC.
}
\label{tab:recombinant_validation}
\centering
\begin{tabular}{llll}
\toprule
Method & Total & QC Pass & QC Fail \\
\midrule
3SEQ & 383 & 71 & 312 \\
CovRecomb & 92& 90 & 2\\
rebar & 213& 203 & 10 \\
\bottomrule
\end{tabular}
\end{table}
